---
title: 'Assignment 6: Data Scraping'
author: "Carly Rauh"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
# load familiar packages
library(tidyverse)
library(rvest)

# check working directory 
getwd()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 

# read url and fetch web resources 
LWSP_webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')
LWSP_webpage

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 

# scraping the data 
Water_system_name <- LWSP_webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
Water_system_name

PWSID <- LWSP_webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
PWSID

Ownership <- LWSP_webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
Ownership

MGD <- LWSP_webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
MGD

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 

# Create a dataframe of maximum daily withdrawals
df_MDW <- data.frame("Month" = c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", 
                                 "Jul", "Nov", "Apr", "Aug", "Dec"),
                     "Year" = rep(2024,12),
                     "Water System Name" = rep(Water_system_name, 12),
                     "PWSID" = rep(PWSID, 12),
                     "Ownership" = rep(Ownership, 12), 
                     "Maximum Day Use" = as.numeric(MGD))

df_MDW <- df_MDW %>% 
  mutate(Date = my(paste(Month,Year)))
                     

#5 

# plot of maximum daily water withdrawal for 2024 by months 
MDW_plot <- ggplot(df_MDW,aes(x=Date,y=as.numeric(MGD))) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("Maximum Daily Water Usage in 2024 for City of Durham"),
       y="Maximum Daily Withdrawl (mgd)",
       x="Date")
MDW_plot


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 

# making scrape function 
scrape <- function(the_PWSID, year){ #getting url
  url <- paste0('https://www.ncwater.org/WUDC/app/LWSP', '/', "report.php?", "pwsid=", the_PWSID, '&', "year=", year)
  print(url)
# calling the website 
  LWSP_website <- read_html(url)
# scraping data
  Water_system_name <- LWSP_website %>% html_nodes("div+ table tr:nth-child(1) 
                                                   td:nth-child(2)") %>% html_text()
  PWSID <- LWSP_website %>% html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
    html_text()
  Ownership <- LWSP_website %>% html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
  MGD <- LWSP_website %>% html_nodes("th~ td+ td") %>% html_text()
# creating dataframe
  the_df <- data.frame("Month" = c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", 
                                 "Jul", "Nov", "Apr", "Aug", "Dec"),
                     "Year" = rep(year,12),
                     "Water System Name" = rep(Water_system_name, 12),
                     "PWSID" = rep(PWSID, 12),
                     "Ownership" = rep(Ownership, 12), 
                     "Maximum Day Use" = as.numeric(MGD)) %>% 
    mutate(Date = my(paste(Month,Year)))
  
  return(the_df)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
# scraping Durham, 2020 data 
df2020 <- scrape('03-32-010',2020)

# plotting max daily withdrawls 
durham_2020_plot <- ggplot(df2020,aes(x=Date,y=Maximum.Day.Use)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("Maximum Daily Water Usage in 2020 for City of Durham"),
       y="Maximum Daily Withdrawl (mgd)",
       x="Date")
  
durham_2020_plot  

# showing structure of dataframe
str(df2020)

```

8. Use the function above to extract data for Cary (PWSID = '03-32-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 

# scraping Cary, 2020 data 
df_cary_2020 <- scrape('03-92-020',2020)

# combining the dataframes 
combine_2020_data <- rbind(df2020, df_cary_2020)

# make combined plot 
combine_2020_plot <- ggplot(combine_2020_data,aes(x=Date,y=Maximum.Day.Use, color=Water.System.Name,group=Water.System.Name)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("Comparing Maximum Daily Water Usage in 2020 for Cities of Durham and Cary, NC"),
       y="Maximum Daily Withdrawl (mgd)",
       x="Date",
       color= "City") +
  theme(legend.position = 'top')

combine_2020_plot

```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 

df_cary_2020 <- scrape('03-92-020',2020)

# subsetting the PWSID
cary_subset <- rep.int('03-92-020', 6)
# listing the years
years <- c(2018,2019,2020,2021,2022,2023)
# mapping the scrape function and getting the data 
cary_df <- map2(cary_subset, years, scrape) 
# combining to a single dataframe
cary_plot <- bind_rows(cary_df)

# plotting the years 
ggplot(cary_plot,aes(x=Month,y=Maximum.Day.Use,group=Year,color=factor(Year))) +
  geom_line() +
  geom_smooth(method="loess",se=FALSE) +
  scale_x_discrete(
    limits = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
  ) +
  labs(title = paste("Cary, NC Maximum Daily Water Usage in 2018-2023"),
       y="Maximum Daily Withdrawl (mgd)",
       x="Month",
       color="Year"
       ) +
      theme(legend.position = 'top')

```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer:
> Yes, Cary appears to have a seasonal trend in water usage over time based on the years we are observing. Water usage tends to increase in the warmer months May-September and decreases in the winter months. Water usage is around 30 mgd in the warmer months but much less in the colder months around 24 mgd or less. Water usage likely increases in the warmer months because of changing human habits in warmer months like taking more showers, watering lawns, and spending more time doing outdoor activities involving water like cleaning or filling up swimming pools. 
